---
title: "Impact of School Choice and Family Engagement on K-12 Educational Success"
subtitle: 'Phase 1: Data Understanding & Preparation'
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    df_print: kable
  word_document:
    toc: true
    toc_depth: '3'
---

\newpage

# Executive Summary
This analysis investigates the impact of school choice and family engagement on K-12 educational success using data from the Parent and Family Involvement (PFI) in Education Survey. Our objective is to develop and evaluate generalized linear models that quantify these relationships, providing actionable recommendations for GVSU's K-12 Connect program. We employ multiple modeling techniques including logistic regression, multinomial regression, discriminant analysis, and Poisson regression to comprehensively understand these educational dynamics.

# Setup and Configuration
## Load all required libraries
```{r}
library(knitr)
library(kableExtra)

library(tidyverse)   # includes dplyr first
library(tidymodels)
library(readxl)
library(ggplot2)
library(corrplot)
library(gridExtra)
library(viridis)
library(scales)
library(readr)

library(nnet)
library(MASS)
library(pscl)
library(car)

# Ensure dplyr wins conflicts
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("mutate", "dplyr")
conflict_prefer("lag", "dplyr")

# Global options for consistent output
options(scipen = 999)  
options(digits = 4)   

# Consistent theme for all plots
theme_set(theme_minimal() + 
          theme(plot.title = element_text(face = "bold", size = 14),
                plot.subtitle = element_text(size = 12),
                axis.title = element_text(face = "bold"),
                legend.position = "bottom",
                panel.grid.minor = element_blank()))

# Color palette for consistent visualizations
project_colors <- c("#2E86AB", "#A23B72", "#F18F01", "#C73E1D", "#6A994E")       
```

# Data Understanding and Preparation
```{r}
data_overview <- data.frame(
  Metric = c("Survey Year", "Total Students", "Variables Analyzed", "Primary Focus Areas"),
  Value = c("2019", "15,500", "71", "School Choice, Family Engagement, Academic Success")
)

kable(data_overview,
      caption = "Table 1: Dataset Overview",
      col.names = c("", ""),
      align = c("l", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "center") %>%
  column_spec(1, bold = TRUE, width = "5cm") %>%
  column_spec(2, width = "6cm")
```


## Project Context and Objectives
GVSU's K-12 Connect has commissioned this analysis to understand how school choice and family engagement in educational activities impact student academic success. The key research objectives are:

1. **Quantify the impact of school choice** on student academic achievement
2. **Measure the effect of family engagement** in both homework support and school activities
3. **Identify optimal engagement patterns** that maximize student success
4. **Provide evidence-based recommendations** for educational interventions

This analysis utilizes data from the U.S. Department of Education's Parent and Family Involvement in Education Survey, which provides comprehensive information about educational experiences, family involvement, and student outcomes.

## Data Loading and Initial Assessment
```{r}
pfi_2016 <- read_csv("pfi-data-2016.csv", show_col_types = FALSE)
pfi_2019 <- read_csv("pfi-data-2019.csv", show_col_types = FALSE)
```

# Handle any initial data type issues
# CSV files sometimes read numeric codes as characters
```{r}
pfi_2016 <- pfi_2016 %>%
  mutate(across(where(is.character), ~na_if(., ""))) %>%
  mutate(across(where(is.character), ~na_if(., "NA")))

pfi_2019 <- pfi_2019 %>%
  mutate(across(where(is.character), ~na_if(., ""))) %>%
  mutate(across(where(is.character), ~na_if(., "NA")))

data_summary <- data.frame(
  Dataset = c("2016 PFI Survey", "2019 PFI Survey"),
  Observations = c(nrow(pfi_2016), nrow(pfi_2019)),
  Variables = c(ncol(pfi_2016), ncol(pfi_2019)),
  Size_MB = c(object.size(pfi_2016)/1048576, object.size(pfi_2019)/1048576)
)

kable(data_summary, 
      caption = "Parent and Family Involvement Survey Data Overview",
      col.names = c("Dataset", "Observations", "Variables", "Size (MB)"),
      align = c("l", "r", "r", "r"),
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                position = "center")

# Compare variable availability across years
vars_2016 <- names(pfi_2016)
vars_2019 <- names(pfi_2019)
common_vars <- intersect(vars_2016, vars_2019)
unique_2016 <- setdiff(vars_2016, vars_2019)
unique_2019 <- setdiff(vars_2019, vars_2016)

cat("\n=== Variable Comparison Across Years ===\n")
cat("Common variables:", length(common_vars), "\n")
cat("Unique to 2016:", length(unique_2016), "\n")
cat("Unique to 2019:", length(unique_2019), "\n")

# Display first few variable names to understand naming convention
cat("\nFirst 10 variables in 2019 dataset:\n")
head(names(pfi_2019), 10)
```
## Data Selection Strategy

Based on our initial assessment, we have made the following strategic decisions:

### Primary Dataset Selection

We selected the **2019 PFI dataset as our primary analysis dataset** for the following reasons:

1. **Recency**: Provides the most current snapshot of educational experiences
2. **Completeness**: Contains all core variables needed for our analysis
3. **Policy Relevance**: Reflects current educational landscape and choice options
4. **Pre-pandemic Data**: Captures typical educational patterns without COVID-19 disruptions

The 2016 dataset will be retained for temporal validation and sensitivity analysis to ensure our findings are stable across time.

```{r}
# Set primary dataset for analysis
pfi_data <- pfi_2019
cat("\nPrimary dataset selected: 2019 PFI Survey\n")
cat("Working with", nrow(pfi_data), "observations and", ncol(pfi_data), "variables\n")

# Quick data quality check
cat("\n=== Initial Data Quality Check ===\n")
cat("Numeric variables:", sum(sapply(pfi_data, is.numeric)), "\n")
cat("Character variables:", sum(sapply(pfi_data, is.character)), "\n")
cat("Logical variables:", sum(sapply(pfi_data, is.logical)), "\n")
```

## Variable Identification and Classification

Our analysis requires careful identification of variables across four critical categories: outcome measures, school choice indicators, family engagement metrics, and demographic controls. We systematically reviewed the PFI codebook and identified relevant variables for each category.
```{r}
## Systematic Variable Identification with Actual Variable Names

# Outcome variables
outcome_vars <- c(
  "SEGRADES",   # Child's grades across all subjects (PRIMARY OUTCOME)
  "SEGRADEQ",   # Expected educational attainment
  "SEENJOY",    # School enjoyment
  "SEABSNT"     # School absences (alternative outcome)
)

# School choice variables - REQUIRED FOCUS AREA
school_choice_vars <- c(
  "SCCHOICE",   # Had choice of schools (KEY VARIABLE)
  "SPUBCHOIX",  # Public school choice
  "SCONSIDR",   # Schools considered
  "EDCPUB",     # Type of school (public/private indicator)
  "SCHLHRSWK",  # School hours per week
  "MOSTIMPT"    # Most important reason for school choice
)

# Family engagement - Homework (REQUIRED FOCUS)
homework_engagement_vars <- c(
  "FHHOME",     # Homework activities at home
  "FHWKHRS",    # Hours of homework per week
  "FHAMOUNT",   # Amount of homework
  "FHCAMT",     # Homework amount appropriate
  "FHPLACE",    # Has place for homework
  "FHCHECKX",   # Check homework completion
  "FHHELP"      # Help with homework
)

# Family engagement - School activities (REQUIRED FOCUS)
school_activity_vars <- c(
  "FSSPORTX",   # School sports participation
  "FSVOL",      # Volunteered at school
  "FSMTNG",     # Attended general school meeting
  "FSPTMTNG",   # Attended parent-teacher conference
  "FSATCNFN",   # Attended school conference
  "FSFUNDRS",   # Participated in fundraising
  "FSCOMMTE",   # Served on school committee
  "FSCOUNSLR",  # Met with school counselor
  "FSFREQ"      # Frequency of school involvement
)

# Parent-school communication variables
communication_vars <- c(
  "FSNOTESX",   # School notes/emails
  "FSMEMO",     # School memos
  "FCSCHOOL",   # Contact with school
  "FCTEACHR",   # Contact with teacher
  "FCSTDS",     # Discuss standards
  "FCORDER",    # Discuss order/discipline
  "FCSUPPRT"    # School support
)

# Home educational activities
home_activity_vars <- c(
  "FOSTORY2X",  # Tell stories to child
  "FOCRAFTS",   # Arts and crafts
  "FOGAMES",    # Play games
  "FOBUILDX",   # Building activities
  "FOSPORT",    # Sports activities
  "FORESPON",   # Discuss responsibilities
  "FOHISTX",    # Discuss history/heritage
  "FODINNERX",  # Family dinners
  "FOLIBRAYX",  # Visit library
  "FOBOOKSTX"   # Visit bookstore
)

# Demographic and control variables
demographic_vars <- c(
  "CDOBMM",     # Child's birth month
  "CDOBYY",     # Child's birth year
  "CSEX",       # Child's sex
  "ALLGRADEX",  # Current grade level
  "CSPEAKX",    # Language spoken
  "HHTOTALXX",  # Household total
  "RELATION",   # Relationship to child
  "P1REL",      # Parent 1 relationship
  "P1SEX",      # Parent 1 sex
  "P1MRSTA",    # Parent 1 marital status
  "P1EMPL",     # Parent 1 employment
  "P1HRSWK",    # Parent 1 hours worked
  "P1AGE",      # Parent 1 age
  "P2GUARD",    # Parent 2 guardian status
  "TTLHHINC",   # Total household income
  "OWNRNTHB",   # Own or rent home
  "CHLDNT",     # Number of children
  "DSBLTY",     # Disability status
  "HHPARN19X",  # Number of parents in household
  "NUMSIBSX",   # Number of siblings
  "PARGRADEX",  # Parent's highest education
  "RACEETH",    # Race/ethnicity
  "CENREG",     # Census region
  "ZIPLOCL"     # ZIP code locale
)

# Additional variables for analysis
other_important_vars <- c(
  "HDHEALTH",   # Child's health
  "SEFUTUREX",  # Future expectations
  "INTACC",     # Internet access
  "EINTNET"     # Internet use for education
)

# Compile all selected variables
all_selected_vars <- unique(c(outcome_vars, school_choice_vars, homework_engagement_vars,
                              school_activity_vars, communication_vars, home_activity_vars, 
                              demographic_vars, other_important_vars))

cat("\n=== Variable Selection Summary ===\n")
cat("Outcome variables identified:", length(outcome_vars), "\n")
cat("School choice variables:", length(school_choice_vars), "\n")
cat("Homework engagement variables:", length(homework_engagement_vars), "\n")
cat("School activity variables:", length(school_activity_vars), "\n")
cat("Communication variables:", length(communication_vars), "\n")
cat("Home activity variables:", length(home_activity_vars), "\n")
cat("Demographic/control variables:", length(demographic_vars), "\n")
cat("Other important variables:", length(other_important_vars), "\n")
cat("Total unique variables selected:", length(all_selected_vars), "\n")
```

```{r}
## Verify Variable Availability and Create Working Dataset

# Check which selected variables are actually in the dataset
available_vars <- intersect(all_selected_vars, names(pfi_data))
missing_vars <- setdiff(all_selected_vars, names(pfi_data))

cat("\n=== Variable Availability Check ===\n")
cat("Variables found in dataset:", length(available_vars), "\n")
cat("Variables not found:", length(missing_vars), "\n")

if(length(missing_vars) > 0) {
  cat("\nVariables not found in dataset:\n")
  print(missing_vars)
}

# Create initial working dataset with available variables
pfi_working <- pfi_data %>%
  dplyr::select(dplyr::all_of(available_vars))

cat("\nWorking dataset created with", ncol(pfi_working), "variables\n")

# Display structure summary
cat("\n=== Working Dataset Structure ===\n")
cat("Observations:", nrow(pfi_working), "\n")
cat("Variables:", ncol(pfi_working), "\n")
cat("Complete cases:", sum(complete.cases(pfi_working)), "\n")
cat("Percentage complete:", round(sum(complete.cases(pfi_working))/nrow(pfi_working)*100, 2), "%\n")
```

```{r}
## Data Type Verification and Key Variable Analysis

# Function to examine variable
examine_variable <- function(data, var_name, var_description = "") {
  if(var_name %in% names(data)) {
    cat(paste0("\n", var_name, " - ", var_description, "\n"))
    cat(paste0(rep("-", 50), collapse=""), "\n")
    
    # Get unique values
    unique_vals <- sort(unique(data[[var_name]]))
    n_unique <- length(unique_vals)
    
    if(n_unique <= 20) {
      cat("Unique values:", paste(unique_vals, collapse = ", "), "\n")
      
      # Show frequency table for categorical-like variables
      freq_table <- table(data[[var_name]], useNA = "ifany")
      prop_table <- prop.table(freq_table) * 100
      
      cat("\nFrequency distribution:\n")
      for(i in 1:length(freq_table)) {
        cat(sprintf("  %s: %d (%.1f%%)\n", 
                   names(freq_table)[i], 
                   freq_table[i], 
                   prop_table[i]))
      }
    } else {
      cat("Continuous variable with", n_unique, "unique values\n")
      cat("\nSummary statistics:\n")
      print(summary(data[[var_name]]))
    }
    
    # Check for missing data codes
    potential_missing <- sum(data[[var_name]] %in% c(-1, -9, 99, 999), na.rm = TRUE)
    if(potential_missing > 0) {
      cat("\nWARNING: Found", potential_missing, "potential missing value codes\n")
    }
  } else {
    cat(paste0("\n", var_name, " - NOT FOUND in dataset\n"))
  }
}

# Examine key outcome variables
cat("\n========== KEY OUTCOME VARIABLES ==========\n")
examine_variable(pfi_working, "SEGRADES", "Student grades (Primary Outcome)")
examine_variable(pfi_working, "SEGRADEQ", "Expected educational attainment")
examine_variable(pfi_working, "SEENJOY", "School enjoyment")

# Examine key school choice variables
cat("\n\n========== KEY SCHOOL CHOICE VARIABLES ==========\n")
examine_variable(pfi_working, "SCCHOICE", "Had choice of schools")
examine_variable(pfi_working, "SPUBCHOIX", "Public school choice")
examine_variable(pfi_working, "EDCPUB", "School type")

# Examine key engagement variables
cat("\n\n========== KEY ENGAGEMENT VARIABLES ==========\n")
examine_variable(pfi_working, "FHHELP", "Homework help")
examine_variable(pfi_working, "FSVOL", "Volunteer at school")
examine_variable(pfi_working, "FSPTMTNG", "Parent-teacher conference")
```

```{r}
## Missing Data Analysis
library(tidyverse)

# Calculate missing data for all variables
missing_summary <- pfi_working %>%
  summarise_all(~sum(is.na(.) | . %in% c(-1, -9, 99, 999))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  mutate(
    Missing_Percent = round(Missing_Count / nrow(pfi_working) * 100, 2),
    Category = case_when(
      Variable %in% outcome_vars ~ "Outcome",
      Variable %in% school_choice_vars ~ "School Choice",
      Variable %in% homework_engagement_vars ~ "Homework",
      Variable %in% school_activity_vars ~ "School Activities",
      Variable %in% communication_vars ~ "Communication",
      Variable %in% home_activity_vars ~ "Home Activities",
      Variable %in% demographic_vars ~ "Demographic",
      TRUE ~ "Other"
    )
  ) %>%
  arrange(Category, Missing_Percent)

# Summary by category
missing_by_category <- missing_summary %>%
  group_by(Category) %>%
  summarise(
    Variables = n(),
    Avg_Missing = round(mean(Missing_Percent), 2),
    Min_Missing = round(min(Missing_Percent), 2),
    Max_Missing = round(max(Missing_Percent), 2)
  ) %>%
  arrange(Avg_Missing)

kable(missing_by_category,
      caption = "Missing Data Summary by Variable Category",
      col.names = c("Category", "# Variables", "Avg % Missing", 
                    "Min % Missing", "Max % Missing")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "center")

# Visualize missing data patterns for key variables
key_vars_missing <- missing_summary %>%
  filter(Variable %in% c(outcome_vars, 
                         c("SCCHOICE", "SPUBCHOIX", "FHHELP", 
                           "FSVOL", "FSPTMTNG", "TTLHHINC"))) %>%
  ggplot(aes(x = reorder(Variable, Missing_Percent), 
             y = Missing_Percent, 
             fill = Category)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = project_colors) +
  labs(
    title = "Missing Data in Key Variables",
    subtitle = "Critical variables for analysis",
    x = "Variable",
    y = "Percentage Missing (%)",
    fill = "Category"
  ) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "red", alpha = 0.5) +
  theme(axis.text.y = element_text(size = 10))

print(key_vars_missing)
```

## Feature Engineering Planning

Based on our variable exploration, we will create several engineered features to better capture the relationships between school choice, family engagement, and academic success.
```{r}
## Feature Engineering Implementation

# Create engineered features
pfi_engineered <- pfi_working %>%
  mutate(
    # Binary academic success
    success_binary = case_when(
      SEGRADES %in% c(1, 2) ~ 1,
      SEGRADES %in% c(3, 4) ~ 0,
      TRUE ~ NA_real_
    ),

    # School activity engagement count
    engagement_count = rowSums(
      dplyr::select(., FSSPORTX, FSVOL, FSMTNG, FSPTMTNG, FSATCNFN,
                    FSFUNDRS, FSCOMMTE, FSCOUNSLR) == 1,
      na.rm = TRUE
    ),

    # Homework intensity index
    homework_intensity = rowMeans(
      dplyr::select(., FHHOME, FHCHECKX, FHHELP),
      na.rm = TRUE
    ),

    # School involvement index
    school_involvement = rowMeans(
      dplyr::select(., FSVOL, FSMTNG, FSPTMTNG, FSCOMMTE),
      na.rm = TRUE
    ),

    # Communication frequency index
    communication_index = rowMeans(
      dplyr::select(., FCSCHOOL, FCTEACHR, FSNOTESX),
      na.rm = TRUE
    ),

    # Home enrichment activities
    enrichment_count = rowSums(
      dplyr::select(., FOSTORY2X, FOCRAFTS, FOGAMES, FOBUILDX,
                    FOSPORT, FOLIBRAYX, FOBOOKSTX) == 1,
      na.rm = TRUE
    ),

    # School choice indicator
    has_choice = ifelse(SCCHOICE == 1, 1, 0),

    # Child age
    child_age = 2019 - as.numeric(CDOBYY)
  )


# Summary of engineered features
engineered_summary <- data.frame(
  Feature = c("success_binary", "engagement_count", "homework_intensity",
              "school_involvement", "communication_index", "enrichment_count"),
  Description = c(
    "Binary outcome: 1 if grades A/B, 0 if C/D",
    "Count of school activities participated in",
    "Average homework engagement (0-1 scale)",
    "Average school involvement (0-1 scale)",
    "Average parent-school communication (0-1 scale)",
    "Count of home enrichment activities"
  ),
  Mean = c(
    mean(pfi_engineered$success_binary, na.rm = TRUE),
    mean(pfi_engineered$engagement_count, na.rm = TRUE),
    mean(pfi_engineered$homework_intensity, na.rm = TRUE),
    mean(pfi_engineered$school_involvement, na.rm = TRUE),
    mean(pfi_engineered$communication_index, na.rm = TRUE),
    mean(pfi_engineered$enrichment_count, na.rm = TRUE)
  ),
  SD = c(
    sd(pfi_engineered$success_binary, na.rm = TRUE),
    sd(pfi_engineered$engagement_count, na.rm = TRUE),
    sd(pfi_engineered$homework_intensity, na.rm = TRUE),
    sd(pfi_engineered$school_involvement, na.rm = TRUE),
    sd(pfi_engineered$communication_index, na.rm = TRUE),
    sd(pfi_engineered$enrichment_count, na.rm = TRUE)
  )
)

kable(engineered_summary,
      caption = "Summary of Engineered Features",
      col.names = c("Feature", "Description", "Mean", "SD"),
      digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "center")

# Visualize distribution of engagement count (for Poisson regression)
engagement_dist <- ggplot(pfi_engineered, aes(x = engagement_count)) +
  geom_histogram(binwidth = 1, fill = project_colors[1], color = "white") +
  labs(
    title = "Distribution of School Activity Engagement",
    subtitle = "Count variable for Poisson regression",
    x = "Number of School Activities",
    y = "Frequency"
  ) +
  scale_x_continuous(breaks = 0:8) +
  theme_minimal()

print(engagement_dist)

# Check for overdispersion in count data
cat("\n=== Poisson Regression Preparation ===\n")
cat("Engagement count statistics:\n")
cat("Mean:", mean(pfi_engineered$engagement_count, na.rm = TRUE), "\n")
cat("Variance:", var(pfi_engineered$engagement_count, na.rm = TRUE), "\n")
cat("Variance/Mean ratio:", 
    var(pfi_engineered$engagement_count, na.rm = TRUE) / 
    mean(pfi_engineered$engagement_count, na.rm = TRUE), "\n")
if(var(pfi_engineered$engagement_count, na.rm = TRUE) > 
   mean(pfi_engineered$engagement_count, na.rm = TRUE)) {
  cat("Note: Variance > Mean suggests overdispersion - consider negative binomial\n")
}
```

## Data Quality Assessment Summary

We have completed a comprehensive assessment of data quality, identifying patterns that will inform our modeling strategy.

```{r}
## Create Comprehensive Data Quality Report

# Overall data quality metrics
quality_report <- list(
  total_obs = nrow(pfi_data),
  working_vars = ncol(pfi_working),
  engineered_vars = ncol(pfi_engineered) - ncol(pfi_working),
  complete_cases = sum(complete.cases(pfi_engineered)),
  complete_pct = round(sum(complete.cases(pfi_engineered))/nrow(pfi_engineered)*100, 2),
  
  # Outcome variable quality
  outcome_missing = round(sum(is.na(pfi_engineered$SEGRADES))/nrow(pfi_engineered)*100, 2),
  binary_outcome_dist = round(prop.table(table(pfi_engineered$success_binary))*100, 2),
  
  # Key predictor quality
  choice_missing = round(sum(is.na(pfi_engineered$SCCHOICE))/nrow(pfi_engineered)*100, 2),
  homework_missing = round(sum(is.na(pfi_engineered$FHHELP))/nrow(pfi_engineered)*100, 2),
  
  # Engagement measures
  mean_engagement = round(mean(pfi_engineered$engagement_count, na.rm = TRUE), 2),
  max_engagement = max(pfi_engineered$engagement_count, na.rm = TRUE)
)

# Create formatted quality report table
quality_df <- data.frame(
  Metric = c("Total Observations", 
             "Working Variables",
             "Engineered Features",
             "Complete Cases",
             "Complete Cases %",
             "Outcome Missing %",
             "Success Rate (Binary)",
             "School Choice Missing %",
             "Homework Help Missing %",
             "Mean Engagement Count",
             "Max Engagement Count"),
  Value = c(quality_report$total_obs,
           quality_report$working_vars,
           quality_report$engineered_vars,
           quality_report$complete_cases,
           paste0(quality_report$complete_pct, "%"),
           paste0(quality_report$outcome_missing, "%"),
           paste0(quality_report$binary_outcome_dist[2], "%"),
           paste0(quality_report$choice_missing, "%"),
           paste0(quality_report$homework_missing, "%"),
           quality_report$mean_engagement,
           quality_report$max_engagement)
)

kable(quality_df,
      caption = "Data Quality Assessment Summary",
      col.names = c("Quality Metric", "Value"),
      align = c("l", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "center")
```

## Key Decisions and Preparation for Modeling

Based on our comprehensive data understanding phase, we have made the following strategic decisions:
```{r}
## Document Final Decisions for Phase 2

decisions <- list(
  "Primary Outcome" = "SEGRADES serves as our primary outcome, with binary version (A/B vs C/D) 
                       for logistic regression and original 4-category for multinomial regression",
  
  "Count Outcome" = "engagement_count (0-8 scale) created from school activity variables 
                     for Poisson regression demonstration",
  
  "School Choice Variables" = "SCCHOICE (had choice) and SPUBCHOIX (public choice) as primary indicators, 
                              supplemented by EDCPUB for school type analysis",
  
  "Homework Engagement" = "FHHELP (help frequency), FHCHECKX (checking), and composite homework_intensity 
                          index capture homework support patterns",
  
  "School Activities" = "Individual activities (FSVOL, FSPTMTNG, etc.) plus school_involvement 
                        index measure school engagement",
  
  "Control Strategy" = "TTLHHINC (income), PARGRADEX (parent education), CSEX, ALLGRADEX, 
                       and RACEETH as key demographic controls",
  
  "Missing Data" = "Variables with <10% missing retained; missing value codes (-1, -9, 99) 
                    converted to NA; imputation planned for modeling phase",
  
  "Interaction Terms" = "Plan to explore SCCHOICE × engagement_count and TTLHHINC × homework_intensity 
                        interactions in Phase 2"
)

# Create formatted decision table
decision_df <- data.frame(
  Decision_Area = names(decisions),
  Strategy = unlist(decisions),
  stringsAsFactors = FALSE
)

kable(decision_df,
      caption = "Strategic Decisions for Modeling Phase",
      col.names = c("Decision Area", "Strategy")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "center") %>%
  column_spec(1, bold = TRUE, width = "3cm") %>%
  column_spec(2, width = "12cm")
```

```{r}
## Save Clean Dataset for Next Phase

# Final data preparation
pfi_clean <- pfi_engineered %>%
  # Replace missing value codes with NA
  mutate(across(everything(), ~ifelse(. %in% c(-1, -9, 99, 999), NA, .)))

# Remove variables with >30% missing
high_missing <- missing_summary %>%
  filter(Missing_Percent > 30) %>%
  pull(Variable)

if(length(high_missing) > 0) {
  cat("\nRemoving", length(high_missing), "variables with >30% missing:\n")
  print(high_missing)
  pfi_clean <- pfi_clean %>%
  dplyr::select(-dplyr::all_of(high_missing))
}

# Save the dataset
write_csv(pfi_clean, "pfi_phase1_clean.csv")
saveRDS(pfi_clean, "pfi_phase1_clean.rds")

# Create final summary
final_summary <- data.frame(
  Stage = c("Original 2019 Data", 
            "Variable Selection", 
            "After Engineering", 
            "Final Clean Dataset"),
  Observations = c(nrow(pfi_data), 
                  nrow(pfi_working), 
                  nrow(pfi_engineered), 
                  nrow(pfi_clean)),
  Variables = c(ncol(pfi_data), 
               ncol(pfi_working), 
               ncol(pfi_engineered), 
               ncol(pfi_clean))
)

kable(final_summary,
      caption = "Data Transformation Summary",
      align = c("l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "center")

cat("\n✓ Phase 1 Complete: Clean dataset saved as 'pfi_phase1_clean.csv' and '.rds'\n")
```


```{r}
engineered_summary_complete <- data.frame(
  Feature = c("success_binary", "engagement_count", "homework_intensity",
              "school_involvement", "communication_index", "enrichment_count",
              "has_choice", "child_age"),
  Description = c(
    "Binary outcome: 1 if grades A/B, 0 if C/D",
    "Count of school activities participated in (0-8)",
    "Average homework engagement frequency",
    "Average school involvement frequency",
    "Average parent-school communication frequency",
    "Count of home enrichment activities (0-7)",
    "Binary indicator: 1 if had school choice, 0 otherwise",
    "Child's age calculated from birth year"
  ),
  Type = c("Outcome", "Composite Index", "Composite Index", 
           "Composite Index", "Composite Index", "Composite Index",
           "Simple Recode", "Simple Calculation"),
  Mean = c(
    round(mean(pfi_engineered$success_binary, na.rm = TRUE), 2),
    round(mean(pfi_engineered$engagement_count, na.rm = TRUE), 2),
    round(mean(pfi_engineered$homework_intensity, na.rm = TRUE), 2),
    round(mean(pfi_engineered$school_involvement, na.rm = TRUE), 2),
    round(mean(pfi_engineered$communication_index, na.rm = TRUE), 2),
    round(mean(pfi_engineered$enrichment_count, na.rm = TRUE), 2),
    round(mean(pfi_engineered$has_choice, na.rm = TRUE), 2),
    round(mean(pfi_engineered$child_age, na.rm = TRUE), 2)
  ),
  SD = c(
    round(sd(pfi_engineered$success_binary, na.rm = TRUE), 2),
    round(sd(pfi_engineered$engagement_count, na.rm = TRUE), 2),
    round(sd(pfi_engineered$homework_intensity, na.rm = TRUE), 2),
    round(sd(pfi_engineered$school_involvement, na.rm = TRUE), 2),
    round(sd(pfi_engineered$communication_index, na.rm = TRUE), 2),
    round(sd(pfi_engineered$enrichment_count, na.rm = TRUE), 2),
    round(sd(pfi_engineered$has_choice, na.rm = TRUE), 2),
    round(sd(pfi_engineered$child_age, na.rm = TRUE), 2)
  )
)

kable(engineered_summary_complete,
      caption = "Table: All Engineered Features",
      col.names = c("Feature", "Description", "Type", "Mean", "SD"),
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  row_spec(1:6, background = "#f0f8ff") 
```

